\chapter{Experimental Design}
\label{ch:experiments}

\section{Overview}

All experiments evaluate the six algorithmic strategies---State-Expanded Dijkstra (B1),
Greedy (B2), DP (B3), Standard A* (B4), RF-A* (B5), and PF-A* (proposed)---on the
variable-purchase model.
Every algorithm receives the same inputs: graph $G$, prices $p$, initial fuel $f_0$, tank
capacity $C$, and efficiency $\eta$.
The primary research question is whether PF-A* achieves the same optimal cost as
Dijkstra/DP/A*/RF-A* while expanding fewer states, and whether all optimal algorithms
outperform Greedy in cost.

All experiments are conducted on Microsoft Azure Standard\_D4s\_v5 virtual machines
(4 vCPUs Intel Xeon Platinum 8370C 2.8\,GHz, 16\,GB DDR4, Ubuntu 22.04\,LTS,
Python~3.11, NetworkX~3.x, OSMnx).
Each algorithm is run for 200 warm-up iterations (discarded) followed by 10,000 timed
runs per configuration.
Runtime is measured with \texttt{time.perf\_counter\_ns()} at the algorithmic level,
excluding I/O.
Statistical significance uses the Wilcoxon signed-rank test ($\alpha=0.05$, Bonferroni
correction for multiple comparisons).

\section{Baselines}

\textbf{State-Expanded Dijkstra (B1)} is the cost lower bound: it exhaustively explores
$G^{*}$ and is guaranteed to find the minimum-cost variable-purchase strategy.
All optimal algorithms (DP, A*, RF-A*, PF-A*) must match this cost exactly; any
divergence flags a correctness bug.

\textbf{Full-tank-only Greedy (B2)} is the cost upper bound: it fills to capacity at every
station regardless of price, representing worst-case uninformed driver behavior.

\textbf{RF-A* (B5)} is the direct algorithmic baseline for PF-A*: both are optimal heuristic
search algorithms with the same worst-case complexity, operating on the same state-expanded
graph under the variable-purchase model.
The sole difference is the heuristic---$h_{\mathrm{RF}}(v)$ is fuel-level agnostic;
$h_{\mathrm{PF}}(v,f)$ is fuel-level aware.
Experiment~5 isolates this difference directly.

\section{Evaluation Metrics}

\begin{table}[ht]
\centering
\caption{Evaluation metrics.}
\label{tab:metrics}
\begin{tabular}{lll}
\toprule
Metric & Unit & Purpose \\
\midrule
Total fuel cost      & THB   & Primary objective; must match B1 for optimal algorithms \\
Runtime              & ms    & Computational efficiency \\
Nodes expanded       & count & Search space explored; key metric for PF-A* vs.\ RF-A* \\
Memory consumption   & MB    & Scalability of $O(VF)$ state space \\
Refueling stops     & count & Practical convenience; tracks partial-fill decisions \\
Sub-optimality gap    & \% vs.\ B1 & Cost excess of Greedy (correctness check for others) \\
\bottomrule
\end{tabular}
\end{table}

\section{Experimental Configurations}

\subsection{Experiment 1: Scalability (Graph Size)}

Fix $\sigma=10\%$, $C=40$\,L, $f_0=0.7C=28$\,L; vary graph size.

\begin{center}
\begin{tabular}{llll}
\toprule
Size & $|V|$ & $|E|$ (approx.) & Runs \\
\midrule
Small  & 100    & 500    & 10,000 \\
Medium & 1,000  & 5,000  & 10,000 \\
Large  & 10,000 & 50,000 & 10,000 \\
\bottomrule
\end{tabular}
\end{center}

\textit{Hypothesis:} Greedy scales as $O(V^{2})$; all other algorithms scale as
$O((EF+VF^{2})\log(VF))$.
PF-A* expands fewer nodes than RF-A* at all sizes, with the gap growing as density
increases.

\subsection{Experiment 2: Fuel Price Variance}

Fix $|V|=1{,}000$, $C=40$\,L, $f_0=28$\,L; vary $\sigma \in \{0\%,5\%,10\%,15\%,20\%\}$.

\textit{Hypothesis:} At $\sigma=0\%$ (uniform prices), all algorithms produce identical costs
and PF-A* offers no node-expansion advantage over RF-A*.
As $\sigma$ increases, optimal algorithms save 5--15\% more cost versus Greedy, and PF-A*
expands progressively fewer states than RF-A* because
$p_{\min}^{\mathrm{reach}}(v,f) \gg p_{\min}$ more often.

\subsection{Experiment 3: Tank Capacity and Initial Fuel}

Fix $|V|=1{,}000$, $\sigma=10\%$; vary $C \in \{20,40,60,80\}$\,L with $f_0=0.7C$.

\textit{Hypothesis:} Smaller $C$ tightens the fuel constraint, increasing the importance of
optimal purchase decisions and widening the cost gap between Greedy and optimal algorithms.
PF-A*'s advantage over RF-A* is largest at small $C$.

\subsection{Experiment 4: Variable Initial Fuel Level}

Fix $|V|=1{,}000$, $C=40$\,L, $\sigma=10\%$; vary
$f_0 \in \{0.25C, 0.50C, 0.75C, 1.0C\}$.

\textit{Hypothesis:} Lower $f_0$ forces earlier mandatory refueling decisions.
PF-A*'s heuristic is most informative at low $f_0$ (where $\max(d_{\mathrm{geo}}/\eta - f, 0)$
is large), so its node-expansion advantage over RF-A* should grow as $f_0$ decreases.

\subsection{Experiment 5: RF-A* vs.\ PF-A* Node Expansions (Primary Novelty Experiment)}

Fix $|V|=1{,}000$; jointly vary $\sigma \in \{5\%,10\%,20\%\}$ and station density
$\in \{0.5, 1, 2, 5\}$ per 100\,km.
Run only RF-A* and PF-A*.
Primary metric: nodes expanded.
Secondary check: both must return identical optimal costs on every instance.

\textit{Hypothesis:} PF-A* expands fewer nodes than RF-A* in all configurations with
$\sigma > 0\%$.
The reduction is largest when station density is high (many stations reachable per state, so
$p_{\min}^{\mathrm{reach}}$ differs most from global $p_{\min}$) and price variance is high.

\subsection{Experiment 6: Real Thai Routes}

Bangkok $\to$ Chiang Mai ($\approx$700\,km), Bangkok $\to$ Phuket ($\approx$900\,km),
Bangkok $\to$ Khon Kaen ($\approx$450\,km).
Real OSM topology, EPPO provincial prices, $f_0=0.7C$, $C=40$\,L, $\eta=10$\,km/L.
Each route evaluated under both purchase models.

\textit{Purpose:} Validates that synthetic benchmark results transfer to real-world conditions.
Directly tests Khuller et al.'s prediction that partial refuelling saves 5--15\% over
full-tank-only on real price data.

\subsection{Experiment 7: Variable Purchase vs.\ Full-Tank-Only Model Comparison}

Fix $|V|=1{,}000$, $f_0=0.7C$; jointly vary $\sigma \in \{0\%,5\%,10\%,20\%\}$ and
$C \in \{20,40,60,80\}$\,L.
Run both models: full-tank-only Dijkstra on $G$ and variable-purchase Dijkstra (B1) on
$G^{*}$.
Primary metric: cost gap (THB and \%).
Secondary metric: runtime overhead of the variable-purchase model.

\textit{Hypothesis 1 (cost):} Variable purchase strictly dominates full-tank-only in cost on
all instances with $\sigma > 0\%$.
The gap grows with $\sigma$ and shrinks with $C$.

\textit{Hypothesis 2 (runtime overhead):} Variable-purchase Dijkstra is slower than
full-tank-only Dijkstra by a factor of $O(F\log F)$ in theory, expected to be 10--40$\times$
in practice for $C=40$\,L, $\varepsilon=1$\,L.
PF-A* is expected to close this gap significantly.

\textit{Hypothesis 3 (tank size interaction):} At $C=20$\,L the variable-purchase model
saves the most; at $C=80$\,L the savings diminish.
This interaction will be visualized as a $\sigma \times C$ heatmap of cost savings.

\section{Correctness Verification}

Before benchmarking, correctness is validated independently:
\begin{itemize}
    \item \textbf{Unit tests:} each algorithm tested on 50 hand-crafted small graphs with
          known optimal solutions, including the running example
          ($f_0=28$\,L, optimal $=216$\,THB).
    \item \textbf{Cross-validation:} on all synthetic instances, Dijkstra (B1) and DP (B3)
          must agree on optimal cost within floating-point tolerance.
    \item \textbf{PF-A* optimality check:} PF-A* cost must equal Dijkstra (B1) cost exactly
          on all instances; any divergence is a heuristic admissibility violation.
    \item \textbf{Feasibility checks:} every returned solution is verified to never drop below
          0\,L fuel at any point along the route.
    \item \textbf{Infeasibility detection:} instances with no valid path are tested to confirm
          all algorithms correctly return ``infeasible.''
\end{itemize}
