\chapter{Conclusion}
\label{ch:conclusion}

This project formulates fuel-aware route optimization on Thailand road networks as a shortest
path problem with explicit refueling decisions, implements six algorithmic strategies, and
provides formal correctness arguments for the proposed Partial-Fill A* heuristic.
The current synthetic evidence already shows that PF-A* is a strong practical method: it
matches the best observed costs returned by the exact baselines in the tested settings while
substantially reducing runtime and node expansions.
The experiments also confirm that the variable-purchase model offers meaningful cost savings
over full-tank-only routing.

Just as importantly, the repository now supports an end-to-end research workflow rather than
an isolated algorithm prototype.
It includes synthetic benchmark generation, automated result analysis, a real-route data
collection pipeline, and a dedicated Experiment~6 execution path for Thai corridors.
That makes the project reproducible and gives the final dissertation a clear empirical
foundation.

The remaining work is sharply defined.
Before the report is archived as a fully complete dissertation, the Standard A* and RF-A*
cost discrepancy should be audited, the strict-profile synthetic rerun should be executed, and
the real Thai route results should be inserted into Chapter~\ref{ch:results}.
Because the document structure in \texttt{ADA\_Final\_Report} already anticipates those
updates, the last stage is now an integration task rather than a rewrite.
