\chapter{Complexity Analysis}
\label{ch:complexity}

All complexity figures are stated for the variable-purchase model on the state-expanded graph
$G^{*} = (S, E^{*})$.
Let $V = |V|$, $E = |E|$, $F = C/\varepsilon$ (discretized fuel levels),
$\Delta = $ max node degree.

\section{State-Space Derivation}

$|S| = O(VF)$ states.
Per state $(v,f)$: $O(\Delta)$ travel relaxations $+$ $O(F)$ refuel relaxations.
Total relaxations: $O(EF + VF^{2})$.
Each relaxation involves one heap operation at $O(\log(VF))$.
This gives the common worst-case bound for all four search-based algorithms.

\begin{table}[ht]
\centering
\caption{Algorithm complexity summary.}
\label{tab:complexity}
\small
\begin{tabular}{lllcc}
\toprule
Algorithm & Time & Space & Optimal? & Variable purchase? \\
\midrule
Dijkstra (no fuel state)       & $O(E\log V)$                   & $O(V{+}E)$ & No  & No \\
Dijkstra (state-expanded, B1)  & $O((EF{+}VF^{2})\log(VF))$    & $O(VF)$    & Yes & Yes \\
Greedy (B2)                    & $O(V^{2})$                      & $O(V)$     & No  & Yes \\
DP na\"{\i}ve (B3)             & $O(V{\cdot}F^{2}{\cdot}\Delta)$ & $O(VF)$    & Yes & Yes \\
DP optimized (B3)              & $O(EF)$                         & $O(VF)$    & Yes & Yes \\
A* state-expanded (B4)         & $O((EF{+}VF^{2})\log(VF))$    & $O(VF)$    & Yes & Yes \\
RF-A* (B5)                     & $O((EF{+}VF^{2})\log(VF))$    & $O(VF)$    & Yes & Yes \\
PF-A* (proposed)               & $O((EF{+}VF^{2})\log(VF))$    & $O(VF)$    & Yes & Yes \\
\bottomrule
\end{tabular}
\end{table}

\section{Why PF-A* Does Not Improve Worst-Case Complexity}

PF-A*'s heuristic $h_{\mathrm{PF}}(v,f)$ is \emph{state-aware}: it depends on both the
current fuel level $f$ and the set of stations reachable from $(v,f)$.
On instances where the fuel-need reduction dominates the reachable-price term, which is the
regime of interest for real Thai corridors, this yields additional pruning.
However, $h_{\mathrm{PF}}$ and $h_{\mathrm{RF}}$ do not admit a universal pointwise
dominance, as discussed in Section~\ref{sec:pfa_correctness}, and in the worst case
(e.g.\ uniform prices, $p_{\min}^{\mathrm{reach}} = p_{\min}$ everywhere), the two
heuristics coincide and PF-A* degenerates to RF-A*.
The worst-case bound is therefore identical.

The improvement is empirical: on networks with high price variance and dense station
coverage, as expected on real Thai routes, the state-awareness of $h_{\mathrm{PF}}$
assigns lower priority to states that RF-A* would expand unnecessarily.
This is quantified in the experiments of Chapter~\ref{ch:experiments}.

\section{Thailand-Scale Estimate}

With $V = 3{,}000$, $E = 15{,}000$, $F = C/\varepsilon = 40/1 = 40$:
$|S| \approx 120{,}000$ states, $\approx 7.2$\,M expanded edges, and $\approx 90$\,M operations, which is feasible
in well under one second on modern hardware.
